master thesis

### 2018-03-21 Paper - Latex/Paper - Skizze - v4.tex
Paper - Anmerkung...
Paper - Methods - Definition of depht and height

  For about 50 years, people have been working on ancestral state reconstruction. The first paper 
    is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state data 
    \cite{Camin1965}. Different methods have been developed and the question is which method is the 
    most suitable for our problem. \\
  % The first methods were only brute force \todo{Quelle, siehe Fitch: Camin and Sokal 1965}. 
  %   Next came a set of parsimony algorithms such as: Fitch-parsimony \cite{Fitch1971}, 
  %   Wagner-parsimony \cite{Swofford1987} ... \todo{weitere?}. \\
  %   With more and more data, there is now the possibility to use more information to calculate the 
  %   probabilities of the ancestral states. In addition to the states of the leafs, algorithms could 
  %   also use branch lengths. The likelihood based algorithms came more in interest. \\

@@

  This paper thus consists of the following points: (1) Find a suitable ancestral state 
    reconstruction method. (2) Perform reconstructing on the Eucaryotic synthesis tree of OTL. \\
  Point 1 consists of a simulation of our data situation and the application of various possible 
    algorithms to these. To evaluate them based on the results. \\
  The one by Louca et al. implemented Sankoff algorithm is the best in our comparisons. Therefore, 
    point 2 consists of reconstructing the ancestral states and predicting the unknown leaf states. 
    And then an evaluation of the results. \\

  \todotext{Definitions of height (min, max, mean), depth of a tree/node (Distanz zur Wurzel vs 
    distanz zum Blatt)}

@@

      For the maximum parsimony analysis of the real data, all transition probabilities are equated.
        However, the used castor package \cite{Louca2017} offers the possibility to enter different 
        transition probabilities.
      \begin{wrapfigure}{r}{0.5\textwidth}
        \begin{center}
          \includegraphics[trim = 0mm 0mm 0mm 15mm, clip, width=0.48\textwidth]{Figures/40-60.png}
        \end{center}
        \caption{60~\% Free-living - 40~\% Parasites \\ red: parasites, blue: free-living, \\ the threshold is at 0.4}
        \label{fig:Beta distribution}
      \end{wrapfigure}

@@

      In the simulation, the influence of the multifurcation and missing data in leaf nodes on the 
        predictive accuracy of the ancestral state reconstruction algorithms is tested. \\
      For the real data, generalized linear models are compared with poisson respectively binomial 
        regression according to their residuals and BICs.
      \anmerkungstext{Erklären, warum und wozu das gemacht wird. Warum BIC und nicht AIC? Erklären,
        was beide bedeuten, wäre sinnvoll. (Thilo)}

@@

      Royer-Carenzi et al. distinguishes two major classes of ancestral state reconstruction methods: \\
      The first is to explain the current state with the least number of state changes between an 
        ancestor and his child, this is called parsimonious. \\
      The other class she presents involves modeling the character evolution as a stochastic process and 
        using the likelihoods to compute the possible ancestral character states. This is generally done 
        with a continuous time Markov model \cite{RoyerCarenzi2013}. \\

      \todo{Pasqualin et al. unterscheiden noch eine weitere Methode: stochastic mapping...} \\

@@

      %---------------------------------------------------------------------------------------------------
      %---------------------------------------------------------------------------------------------------
      \subsection{Fitch maximum parsimony}
        Fitch maximum parsimony is an algorithm for rooted, binary trees and describes an ancestral state 
          reconstruction for discrete states \cite{Fitch1971} by minimizing transitions between states. \\
        Note, the original Fitch algorithm has the sole purpose of minimizing the number of transitions 
          and not reconstructing the ancestral nodes. Felsenstein \cite{Felsenstein2003} describes a 
          simple extension for the reconstruction. Cunningham et al. \cite{Cunningham1998} have refined 
          these. \todotext{Wir haben mit ein paar kleinen änderungen optimiert... und schließlich auf 
          multifurcated angepasst...} \todotext{eigentlich ist Cunningham 'nur' eine kritische 
          Neubewertung. Sie beziehen ihren Algortihmus auf Swofford und Maddison...} \\
        To understand the differences to the multifurcated case, the algorithm for the binary case is 
          briefly explained and referred to the extension. \\
          \begin{wrapfigure}[19]{r}{0.5\textwidth}
            \centering
            \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
            \caption{Fitch algorithm for binary trees. \\
              The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
              root node) consists of three sets, where the last set is the final one (bold). \\
              From the second internal node (seen from the root node) there are several possibilities to 
              create the second and third set.}
            \label{fig: binary Fitch}
          \end{wrapfigure}
        Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
          a set of states. There are only two states in this paper, free-living (FL) and parasitic (P). 
          Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
          has only one. Leaf nodes have their state as a set (eg \{FL\} or \{P\}, unknown leaf nodes the 
          union of all possible states (\{FL, P\}). \\
        The algorithm traverses three times trough the tree and fills these sets. \\
        In each step, two sets are considered and their intersect formed. There are two cases:
        \begin{enumerate}
          \item The intersection is not empty and corresponds to the new set.
          \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
        \end{enumerate}
        First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal. 
          Each internal node is formed from its child nodes, where at the beginning the only information
          lies. \\
        Second traverse from the root node to the leafs. Each internal node is formed from its father node 
          and its sibling node. \\
        Last traversion (direction does not matter): Build the final state for every node. It is formed 
          from the sets of previous traversals. \\
        (The original Fitch algorithm was designed to minimize transitions without predicting actual states 
          of internal nodes, so it was just the first traversal.) \\
        The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
          more than two children may be present for the first traversal, but the incision or union may 
          also be formed over more than two sets. Also in the second traversing, there may be several 
          sibling nodes. However, there are several possibilities here that were all tested and compared 
          in the simulation. Some of these options are already available in the binary case:
        \begin{itemize}
          \item The father node has (except for the root node) two state sets, because he came through 
            the up-traversing previously. Are both sets used or only the first traversing?
          \item Since there are several siblings, do \colorbox{red}{you} first of all make the cut or union, 
            or directly in the whole with the father node?
        \end{itemize}
        The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
          both possibilities of the three sets. \\
        Cunningham uses only the first state set of the father node \cite{Cunningham1998}. \\
        From these two points four different versions of Fitch were formed:
        \begin{enumerate}
          \item Fitch 1: First state set of father node; intersection/union of siblings first.
          \item Fitch 2: First state set of father node; intersection/union of siblings together with father node.
          \item Fitch 3: Both state sets of father node; intersection/union of siblings first.
          \item Fitch 4: Both state sets of father node; intersection/union of siblings together with father node sets.
        \end{enumerate}
        \begin{figure}
          \centering
          \includegraphics[width=0.8\textwidth]{Figures/simulation_fitch_evaluation.png}
          \caption{Test of Fitch Versions.}
          \label{fig:Fitch versions}
        \end{figure}
        These four versions were tested in the simulation with 100 trees and 10000 leaf nodes and a
          distribution of 60~\% FL to 40~\% P. Figure \ref{fig:Fitch versions} shows this over all unknown 
          node percentage. \\
        At 90~\% unknown nodes and 90~\% of multifurcation of the internal nodes, version 1 was 89.26~\%, 
          version 2 was 89.26~\%, version 3 was 89.35~\%, and version 4 was 89.31~\% correct. Therefore, 
          only version 3 was used for all further simulations.

>>>>

  For about 50 years, people have been working on ancestral state reconstruction. The first paper 
    is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state data 
    \cite{Camin1965}. Different methods have been developed and the question is which method is the 
    most suitable for our problem. \\
  \anmerkungstext{hier mehr dazu? Likelihood vs parsiomony... das befindet sich momentan im Methoden 
    Teil: Ancestral state reconstruction methods} \\
  % The first methods were only brute force \todo{Quelle, siehe Fitch: Camin and Sokal 1965}. 
  %   Next came a set of parsimony algorithms such as: Fitch-parsimony \cite{Fitch1971}, 
  %   Wagner-parsimony \cite{Swofford1987} ... \todo{weitere?}. \\
  %   With more and more data, there is now the possibility to use more information to calculate the 
  %   probabilities of the ancestral states. In addition to the states of the leafs, algorithms could 
  %   also use branch lengths. The likelihood based algorithms came more in interest. \\

@@

  This paper thus consists of the following points: (1) Find a suitable ancestral state 
    reconstruction method. (2) Perform reconstructing on the Eucaryotic synthesis tree of OTL. \\
  Point 1 consists of a simulation of our data situation and the application of various possible 
    algorithms to these. To evaluate them based on the results. \\
  The one by Louca et al. implemented Sankoff algorithm is the best in our comparisons. Therefore, 
    point 2 consists of reconstructing the ancestral states and predicting the unknown leaf states. 
    And then an evaluation of the results.

@@

      For the maximum parsimony analysis of the real data, all transition probabilities are equated.
        However, the used castor package \cite{Louca2017} offers the possibility to enter different 
        transition probabilities.
      \begin{wrapfigure}{r}{0.5\textwidth}
        \begin{center}
          \includegraphics[trim = 0mm 0mm 0mm 13mm, clip, width=0.48\textwidth]{Figures/40-60.png}
        \end{center}
        \caption{60~\% Free-living - 40~\% Parasites \\ red: parasites, blue: free-living, \\ the threshold is at 0.4}
        \label{fig:Beta distribution}
      \end{wrapfigure}

@@

      In the simulation, the influence of the multifurcation and missing data in leaf nodes on the 
        predictive accuracy of the ancestral state reconstruction algorithms is tested. \\
      For the real data, generalized linear models are compared with poisson respectively binomial 
        regression according to their residuals and BICs. \\
      \anmerkungstext{Erklären, warum und wozu das gemacht wird. Warum BIC und nicht AIC? Erklären,
        was beide bedeuten, wäre sinnvoll. (Thilo)} \\

      For each node, depth, min, max and mean height were noted. Where the depth of a node is the 
        distance (number of edges) to the root node and the height of a knot is described as the 
        largest distance to a leaf node. In this work, a distinction is made between minimum, maximum 
        and average distance ($\rightarrow$ min, max and mean height). \\
      The influence in the modeling of these parameters was tested, additive as well as multiplicative.

@@

      Royer-Carenzi et al. distinguishes two major classes of ancestral state reconstruction methods: \\
      The first is to explain the current state with the least number of state changes between an 
        ancestor and his child, this is called parsimonious. \\
      The other class she presents involves modeling the character evolution as a stochastic process and 
        using the likelihoods to compute the possible ancestral character states. This is generally done 
        with a continuous time Markov model \cite{RoyerCarenzi2013}. \\
      \todo{Pasqualin et al. unterscheiden noch eine weitere Methode: stochastic mapping...} \\

@@

      %---------------------------------------------------------------------------------------------------
      %---------------------------------------------------------------------------------------------------
      \subsection{Fitch maximum parsimony}
        Fitch maximum parsimony is an algorithm for rooted, binary trees and describes an ancestral state 
          reconstruction for discrete states \cite{Fitch1971} by minimizing transitions between states. \\
        Note, the original Fitch algorithm has the sole purpose of minimizing the number of transitions 
          and not reconstructing the ancestral nodes. Felsenstein \cite{Felsenstein2003} describes a 
          simple extension for the reconstruction. Cunningham et al. \cite{Cunningham1998} have refined 
          these. \todotext{Wir haben mit ein paar kleinen änderungen optimiert... und schließlich auf 
          multifurcated angepasst...} \todotext{eigentlich ist Cunningham 'nur' eine kritische 
          Neubewertung. Sie beziehen ihren Algortihmus auf Swofford und Maddison...} \\
        To understand the differences to the multifurcated case, the algorithm for the binary case is 
          briefly explained and referred to the extension. \\

        Input: A rooted, binary tree, with state informations in the leaf nodes. Each node is depicted as 
          a set of states. There are only two states in this paper, free-living (FL) and parasitic (P). 
          Internal nodes have three sets, which are empty at the beginning, excluding the root node, it 
          has only one. Leaf nodes have their state as a set (eg \{FL\} or \{P\}, unknown leaf nodes the 
          union of all possible states (\{FL, P\}). \\
        The algorithm traverses three times trough the tree and fills these sets. \\
        In each step, two sets are considered and their intersect formed. There are two cases:
        \begin{enumerate}
          \item The intersection is not empty and corresponds to the new set.
          \item The intersection is empty. $\rightarrow$ Build the union of these sets as new set.
        \end{enumerate}

        First traverse from the leaf nodes to the root / move down the tree / postorder tree traversal.
          \begin{wrapfigure}{r}{0.5\textwidth}
            \centering
            \includegraphics[width=0.4\textwidth]{Figures/Fitch1.png}
            \caption{Fitch algorithm for binary trees. \\
              The unknown leaf node is discribed with both states. Computed internal nodes (exclusive the 
              root node) consists of three sets, where the last set is the final one (bold). \\
              From the second internal node (seen from the root node) there are several possibilities to 
              create the second and third set.}
            \label{fig: binary Fitch}
          \end{wrapfigure} 
          Each internal node is formed from its child nodes, where at the beginning the only information
          lies. \\
        Second traverse from the root node to the leafs. Each internal node is formed from its father node 
          and its sibling node. \\
        Last traversion (direction does not matter): Build the final state for every node. It is formed 
          from the sets of previous traversals. \\
        (The original Fitch algorithm was designed to minimize transitions without predicting actual states 
          of internal nodes, so it was just the first traversal.) \\
        The extension to the non-binary case is quite obvious, but holds some opportunities. In this case, 
          more than two children may be present for the first traversal, but the incision or union may 
          also be formed over more than two sets. Also in the second traversing, there may be several 
          sibling nodes. However, there are several possibilities here that were all tested and compared 
          in the simulation. Some of these options are already available in the binary case:
        \begin{itemize}
          \item The father node has (except for the root node) two state sets, because he came through 
            the up-traversing previously. Are both sets used or only the first traversing?
          \item Since there are several siblings, do \colorbox{red}{you} first of all make the cut or union, 
            or directly in the whole with the father node?
        \end{itemize}
        The first point already has an effect on the binary case. Figure \ref{fig: binary Fitch} shows 
          both possibilities of the three sets. \\
        Cunningham uses only the first state set of the father node \cite{Cunningham1998}. \\
        From these two points four different versions of Fitch were formed:
        \begin{enumerate}
          \item Fitch 1: First state set of father node; intersection/union of siblings first.
          \item Fitch 2: First state set of father node; intersection/union of siblings together with father node.
          \item Fitch 3: Both state sets of father node; intersection/union of siblings first.
          \item Fitch 4: Both state sets of father node; intersection/union of siblings together with father node sets.
        \end{enumerate}
        \begin{figure}
          \centering
          \includegraphics[width=0.8\textwidth]{Figures/simulation_fitch_evaluation.png}
          \caption{Test of Fitch Versions.}
          \label{fig:Fitch versions}
        \end{figure}
        These four versions were tested in the simulation with 100 trees and 10000 leaf nodes and a
          distribution of 60~\% FL to 40~\% P. Figure \ref{fig:Fitch versions} shows this over all unknown 
          node percentage. \\
        At 90~\% unknown nodes and 90~\% of multifurcation of the internal nodes, version 1 was 89.26~\%, 
          version 2 was 89.26~\%, version 3 was 89.35~\%, and version 4 was 89.31~\% correct. Therefore, 
          only version 3 was used for all further simulations.
